### 2024-11-23 Article/Conclusions.tex
minima

In this work, we have defined graphs with equivalences and e-graphs in a more abstract manner than \cite{egg}, independent of specific implementations. This allowe us  to analyze the categorical properties of these structures. Building upon this framework, we have property, in particular, the adhesivivty of graphs with equivalences and e-graphs with respect to the class of regular monomorphisms. This result is potentially useful for advancing state-of-the-art techniques, such as the DPO approach for concurrent updates to the structure, which has not yet been explored.

Both graphs with equivalences and e-graphs can be realized as functors with values in $\Set$. One might ask whether it is strictly necessary to use $\Set$ as the codomain of these functors. Our framework is developed from a more general perspective, using minimal set-theoretic concepts and instead relying on categorical constructions. It only requires $\mathcal{M}$-adhesivity and exactness  \cite{barrexact}. This propert certainly hold in $\Set$ (and more generally every elementary topos \cite{maclane2012sheaves}). Roughly speaking, this is  the property that enabled the equivalence noted in \Cref{rem:eqgrph_set_eq}. Therefore, a potential direction for future work will be to provide a more general presentation of these structures, replacing $\Set$ with any adhesive and exact category.

>>>>

In this work, we have defined graphs with equivalences and e-graphs in an abstract manner, making it independent of specific implementations such as in \cite{egg}. This allows  to analyse the categorical properties of these structures. Building upon this framework, we have proved, in particular, the adhesivity of graphs with equivalences and e-graphs with respect to the class of regular monomorphisms. This result is potentially useful for advancing state-of-the-art techniques, such as the DPO approach for concurrent updates to the structure, which has not yet been explored.

Both graphs with equivalences and e-graphs can be realized as functors with values in $\Set$. One might ask whether it is strictly necessary to use $\Set$ as the codomain of these functors. Our framework is developed from a more general perspective, using minimal set-theoretic concepts and instead relying on categorical constructions. It only requires $\mathcal{M}$-adhesivity and exactness  \cite{barrexact}. These properties certainly holdy in $\Set$ and more generally in every elementary topos \cite{maclane2012sheaves}. Roughly speaking, this is  the property that enabled the equivalence noted in \Cref{rem:eqgrph_set_eq}. Therefore, a potential direction for future work will be to provide a more general presentation of these structures, replacing $\Set$ with any adhesive and exact category.

### 2024-11-23 Article/introduction.tex
minima

The utility of e-graphs thus extends beyond theoretical applications, finding practical relevance in fields such as system optimisation and component upgrading.
%\todo{Qua servirebbero degli esempi di uso}
%
However, the more a formalism is used, the greater is the need for formal frameworks to guide its development and application. In particular, addressing issues of parallelism and concurrency is critical for 
a mathematically 
sound understanding of these structures.

Among the rewriting paradigms for graph-like structures, one of the leading proposal is the Double Pushout (DPO) approach \cite{Ehrig_Golas,ehrig2006fundamentals}, since it provides, paired with the framework of $\mathcal{M}$-adhesive  categories \cite{lack2005adhesive,azzi2019essence,behr2022fundamentals}, a robust mathematical foundation for the definition and the application of rules. 
%
Integrating the theory of e-graphs manipulation into the broader subject of DPO rewriting would put on firmer ground the status of e-graphs. Moreover, it would allow to apply to this new context the well known results about concurrency and parallelism already known provided by the abstract theory of DPO rewriting systems \cite{baldan1999concurrent,ehrig1999handbook}.

Such an integration is the main aim of the thesis.
 On the technical level, to reach our aim we need to prove some adhesivity property of e-graphs. 
 In order to do so, some consideration on graphs with equivalences are in order. 
 A graph with equivalence is nothing but a graph endowed with an equivalence relation among its nodes. 
 The main result on this kind of graphs is a weak adhesivity property that holds for regular monomorphisms. 
% Such result is, then, inherited by e-graphs \todo{Da qualche parte mettere riferimento a \cite{egg}}, which are graphs with equivalence which satisfies additional requirements that can be intuitively summarized as the property of edges to have equivalent sources whenever targets are equivalent.
%
%%On the technical level, to reach our objective we need to prove that the category of e-graphs exhibits a specific adhesivity property. To this end, it is necessary to first examine \emph{graphs with equivalence}, as they form the foundational structure underlying e-graphs.
%
%A graph with equivalence is simply a graph where an equivalence relation is defined on its nodes. The primary result concerning graphs with equivalences is an adhesivity property with respect to regular monomorphisms. 
%
This adhesivity result, in turn, extends naturally, via the results of \cite{castelnovo2022newcriterionmathcalmmathcalnadhesivity}, to e-graphs \todo{Add reference to \cite{egg} da qualche parte}, which in fact can be interpreted as a specialised class of graphs with equivalences that satisfy additional structural constraints. Specifically, e-graphs enforce a condition that edges must respect equivalence relations: if two edges have equivalent target nodes, then their source nodes must also be equivalent.

@@

\smallskip \emph{Synopsis.} \Cref{chap:one} introduces the basic concepts of category theory that will be used throughout this thesis. The main goal of this chapter is to build the categorical toolbox necessary for the remainder of the work. Category theory provides a general and uniform formalism, which has the side effect of making many definitions seem quite abstract. To address this, we will provide concrete examples to help clarify and illustrate the concepts.
\Cref{chap:two} focuses on graphs. The structure we will use to model graphs are presheaves, which not only provides a straightforward formalisation, putting in evidence the categorical properties of graphs that we will need.  We will then endow graphs with an equivalence relation among vertices, and proof that the resulting category satisfies an adhesivity property. Finally, in the last part we will examine
 e-graphs, showing that they inherit the above adhesivity property from graphs with equivalences.

>>>>

The utility of e-graphs thus extends beyond theoretical applications, finding practical relevance in fields such as system optimisation and upgrading.
%\todo{Qua servirebbero degli esempi di uso}
%
However, the more a formalism is used, the greater is the need for formal frameworks to guide its development and application. In particular, addressing issues of parallelism and concurrency is critical for 
a mathematically 
sound understanding of these structures.

Among the rewriting paradigms for graph-like structures, one of the leading proposal is the Double Pushout (DPO) approach \cite{Ehrig_Golas,ehrig2006fundamentals}, since it provides, paired with the framework of $\mathcal{M}$-adhesive  categories \cite{lack2005adhesive,azzi2019essence,behr2022fundamentals}, a robust mathematical foundation for the definition and the application of rules. 
%
Integrating the theory of e-graphs manipulation into the broader subject of DPO rewriting would put on firmer ground the status of e-graphs. Moreover, it would allow to apply to this new context well-known results about concurrency and parallelism already provided by the theory of DPO rewriting systems \cite{baldan1999concurrent,ehrig1999handbook}.

Such an integration is the main aim of the thesis.
 On the technical level, to reach our aim we need to prove some adhesivity property of e-graphs. 
 In order to do so, some considerations on graphs with equivalences are in order. 
 A graph with equivalence is nothing but a graph endowed with an equivalence relation among its nodes. 
 The main result on this kind of graphs is a weak adhesivity property that holds for regular monomorphisms. 
% Such result is, then, inherited by e-graphs \todo{Da qualche parte mettere riferimento a \cite{egg}}, which are graphs with equivalence which satisfies additional requirements that can be intuitively summarized as the property of edges to have equivalent sources whenever targets are equivalent.
%
%%On the technical level, to reach our objective we need to prove that the category of e-graphs exhibits a specific adhesivity property. To this end, it is necessary to first examine \emph{graphs with equivalence}, as they form the foundational structure underlying e-graphs.
%
%A graph with equivalence is simply a graph where an equivalence relation is defined on its nodes. The primary result concerning graphs with equivalences is an adhesivity property with respect to regular monomorphisms. 
%
This adhesivity result, in turn, extends naturally, via the results of \cite{castelnovo2022newcriterionmathcalmmathcalnadhesivity}, to e-graphs \todo{Add reference to \cite{egg} da qualche parte}, which in fact can be interpreted as a specialised class of graphs with equivalences that satisfy additional structural constraints. Specifically, e-graphs enforce a condition that edges must respect equivalence relations: if two edges have equivalent target nodes, then their source nodes must also be equivalent.

@@

\smallskip \emph{Synopsis.} \Cref{chap:one} introduces the basic concepts of category theory used throughout this thesis. The main goal of this chapter is to build the toolbox necessary for the remainder of the work. Category theory provides a general and uniform formalism, which sometimes can make definitions seem quite abstract. To address this, we provide concrete examples to help clarify and illustrate the concepts.
\Cref{chap:two} focuses on graphs. The structure we  use to model graphs are presheaves, which not only provides a straightforward formalisation, putting in evidence the categorical properties of graphs that we  need.  We then endow graphs with an equivalence relation among vertices, and proof that the resulting category satisfies an adhesivity property. Finally, in the last part we examine
 e-graphs, showing that they inherit the above adhesivity property from graphs with equivalences.
